---
title: "p8105_hw3_lz2950"
author: "Lehan Zou"
date: "2023-10-14"
output: github_document
---
```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(dplyr)
library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```



### Problem 0

This solution focuses on a reproducible report containing code and text necessary for Problems 1-3, and is organized as an R Project. This was not prepared as a GitHub repo; examples for repository structure and git commits should be familiar from other elements of the course.

Throughout, we use appropriate text to describe our code and results, and use clear styling to ensure code is readable. 

### Problem 1

#### Read in the data

```{r}
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```

#### Answer questions about the data

This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart |> select(product_id) |> distinct() |> count()` products found in `r instacart |> select(user_id, order_id) |> distinct() |> count()` orders from `r instacart |> select(user_id) |> distinct() |> count()` distinct users.

Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

Next is a plot that shows the number of items ordered in each aisle. Here, aisles are ordered by ascending number of items.

```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

Our next table shows the three most popular items in aisles `baking ingredients`, `dog food care`, and `packaged vegetables fruits`, and includes the number of times each item is ordered in your table.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week. This table has been formatted in an untidy manner for human readers. Pink Lady Apples are generally purchased slightly earlier in the day than Coffee Ice Cream, with the exception of day 5.

```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```


## Problem 2
* data cleaning
```{r}
data("brfss_smart2010")

brfss_data <- brfss_smart2010 |>
  janitor::clean_names()|>
  rename(state = locationabbr, county = locationdesc)|>
  filter(
    topic == "Overall Health",
    response %in% c('Poor', 'Fair', 'Good', 'Very good', 'Excellent')
  )|>
  mutate(
    response = factor(response, levels=c('Poor', 'Fair', 'Good', 'Very good', 'Excellent'))) |>
  arrange(response)
```
* In 2002, which states were observed at 7 or more locations? What about in 2010?
```{r}
states_2002 = brfss_data |>
  filter(year == 2002) |>
  group_by(state) |>
  summarize(
    n = n_distinct(county)
  )|>
  filter(n >= 7 )

states_2010 = brfss_data |>
  filter(year == 2010) |>
  group_by(state) |>
  summarize(
    n = n_distinct(county)
  )|>
  filter(n >= 7 )
```

In 2002, `r pull(states_2002, state)` were observed at 7 or more locations. In 2010, `r pull(states_2010, state)` were observed at 7 or more locations.

* Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state.
```{r}
excellent = brfss_data |>
  filter(response == "Excellent") |>
  group_by(year, state) |>
  summarise(ave_value = mean(data_value, na.rm = T))
```
The resulting data set contains `r nrow(excellent)` observations and `r ncol(excellent)` variables.

* Make a “spaghetti” plot of this average value over time within a state.
```{r}
excellent |>
  ggplot(aes(x = year, y = ave_value, color = state)) + 
  geom_line() +
  labs(title = "Average Data Value of States")
```
This line plot using different color from purple to yellow indicating different states and showing their average data value over time. 


* Finally, make a two-panel plot showing, for the years 2006, and 2010, distribution of `data_value` for responses (“Poor” to “Excellent”) among locations in NY State.
```{r}
brfss_data |>
  filter(state == "NY" & year %in% c(2006, 2010))|>
  ggplot(aes(x = response, y = data_value, color = year)) +
  geom_boxplot() +
  facet_grid(year ~ .) +
  labs(title = "Distribution of Data Value for Responses Among locations in 2006 and 2010",
       x = "Response",
       y = "Data Value")
```
The box plot shows data values in year 2006 and 2010 follows similar distribution while the value in 2010 of each response are higher than 2006. Yet they share the same trend with the value increasing from "Poor" to "Very good" but decreasing from "Very good" to "Excellent". 

## Problem 3

* Load, tidy, merge, and otherwise organize the data sets.
```{r}
edu_mapping <- c("Less than high school","High school equivalent", "More than high school")
sex_mapping <- c("male", "female")

covar = 
  read_csv("data/nhanes_covar.csv", skip = 4) |> 
  janitor::clean_names() |> 
  filter(age >= 21) |> #exclude participants less than 21 years of age
  drop_na() |> #remove NA values
  #encode data with reasonable variable classes 
  mutate(
      education = factor(edu_mapping[education], levels = edu_mapping),
      sex = factor(sex_mapping[sex], levels = sex_mapping)
  )


accel = 
  read_csv("data/nhanes_accel.csv") |>
  janitor::clean_names() |> 
  pivot_longer(
    min1:min1440,
    names_to = "minute",
    values_to = "MIMS_Value",
    names_prefix = "min")
    
    
nhanes_merge = 
  left_join(covar, accel, by = "seqn")
```
The merged dataset contains `r nrow(nhanes_merge)` observations and `r ncol(nhanes_merge)` variables related to 1440 minutes of MIMS value of `r nhanes_merge |> pull(seqn) |> unique() |> length()` participants.

* Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items
```{r}
covar |> 
  count(education, sex) |> 
  pivot_wider(
    names_from = sex,
    values_from = n
  ) |> 
  knitr::kable()
  
```
From the table we clearly see that Less than high school and higher than high school education history among women and men are about the same while there are less women under high school equivalent than men.


* Create a visualization of the age distributions for men and women in each education category.

```{r}
nhanes_merge |> 
  ggplot(aes(x = education, y = age, fill = sex)) +
  geom_boxplot() +
  labs(
    title = "Age distribution for men and women in each education category",
    x = "Education",
    y = "Age")
```
* Using the tidied data set, aggregate across minutes to create a total activity variable for each participant. 

```{r}
activity = 
  nhanes_merge |> 
  group_by(seqn, sex, age, education) |> 
  summarise(activity = sum(MIMS_Value))
```

Plot
```{r}
activity |> 
  ggplot(aes(x = age, y = activity, color = sex)) +
  geom_point() +
  geom_smooth() +
  facet_grid(education~.) +
  labs(title = "Daily MIMS Activity among Ages")
```
The plot shows activity score decreases for both men and women as age increases for all educational level. More specifically, female has higher activity score than men in high school equivalent and more than high school educational level.

* Finally, make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex.
```{r}
nhanes_merge|> 
  ggplot(aes(x = minute, y = MIMS_Value, color = sex)) +
  geom_point() +
  geom_smooth() +
  facet_grid(education ~ .) +
  theme_minimal()+
  labs(title = "MIMS Activity per Minute")
```
This plot shows similar trend for both female and male among different educational level.


